Deciding Bitvector Termination with Program Synthesis

### 2014-07-05 papers/termination/paper.tex
Motivational examples, bibtex

One difficulty with a Ramsey-based approach is that validity of the termination argument can only be checked by considering the \emph{transitive closure}
of the program, rather than a single step.
As such, there is experimental evidence that most of the time is spent in reachability analysis \cite{DBLP:conf/pldi/CookPR06}, 
requiring the support of powerful safety checkers.
Basically, these approaches opt for simpler termination arguments in the detriment of complex validity checking.

As Ramsey-based approaches are limited by the state of the art in safety checking, 
recent research shifts back to more complex termination arguments that are easier to check \cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
%
%
Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
we investigate its extreme: \emph{unrestricted} termination arguments. 
This means that our ranking functions may involve non-linearity and lexicographic orders (we do not commit to any such form, i.e. we do not use templates).

While the summary  in Figure~\ref{fig:handletable} supports the observation that the majority of existing termination analyses are designed for
linear programs and linear ranking functions, it also identifies another common assumption made by state of the art termination provers: 
\emph{the equivalence of bit-vector/float semantics and integer/real semantics}. Thus, most of the techniques treat
bit-vectors and floats as mathematical integers and reals, respectively  \cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

By assuming bit-vector/float semantics to be equivalent to integer/real semantics, 
these techniques ignore the wrap-around behaviour caused by under/over-flows as well as width conversions, 
and ultimately diverge from the execution of a program on a computer.  
In Section~\ref{sec:motivation}, we show that integers/reals and bit-vectors/floats exhibit
incomparable behaviours with respect to termination, e.g.
programs that terminate for integers may \emph{not} terminate for bit-vectors.
Thus, abstracting bit-vectors/floats to integers/reals may give rise to 
{\em unsound} and {\em incomplete} analyses.
%design decisions
%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.

@@

We propose a general framework that %does not assume the existence of linear ranking functions and can 
uniformly computes
\emph{lexicographic/non-lexicographic} \emph{linear/non-linear} 
ranking functions supported by inductive \emph{linear/non-linear} invariants for loops with 
\emph{linear/non-linear} guards and transitions over bit-vectors and floats.
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument. 
We combine the generation of ranking functions with the generation of invariants such that the necessary 
supporting invariants and the ranking function are discovered at the same time.

For this purpose, we phrase the termination and non-termination problems as second order satisfaction problem and
we identify a fragment of second-order logic that is expressive enough to capture our formulation, 
which we call \emph{the synthesis fragment}. Subsequently, we propose a method 
for checking the satisfiability of a formula in the synthesis fragment as program synthesis (Section~\ref{sec:synthesis}). 
Our method is theoretically complete, i.e. if a pair of a ranking function and a supporting invariant 
exist, it will be found.

With respect to the performance of our technique, we show that its runtime is dominated by the size of 
the shortest termination proof.
%While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors 
%influencing our method's performance and
%find that our technique  
This means that our procedure is not directly dependent on the structure of the analysed loop, e.g. number of variable, number of lassos, 
but on the Kolmogorov complexity of its minimal termination argument. 
\emph{In our design, proving a program terminates becomes as hard as finding its minimal temination argument}.
We argue that the length of a program's termination proof is indicative of how easy to understand the program is, and thus 
programmers tend to write programs that have relatively short termination proofs. 
We empirically show that this claim holds in practice, causing our technique to perform well.

@@

@@

Figure~\ref{fig:handletable} illustrates the most common assumptions made by existent termination analyses:
\begin{itemize}
\item[(i)] programs are linear.
\item[(ii)] terminating programs have linear termination arguments.
\item[(iii)] bit-vectors semantics and mathematical integers semantics are equivalent.
\item[(iv)] floats semantics and mathematical reals semantics are equivalent.
\end{itemize}

%In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}). 
Next, we show how these assumptions are broken by even simple programs that cannot therefore be handled accurately by state of the art termination tools.

@@

%Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.
The loop in Figure~\ref{fig:motivation.a} breaks assumption (i) as it makes use of the bit-wise $\&$ operator.
%As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
%although a linear ranking function does exist (see Figure~\ref{fig:handletable}). 
%In order to find a ranking function for this example, it is necessary to take into account
%the semantics of the bit-wise AND operator.
Our technique finds that a possible ranking function is the linear function
$R(x) = x$, whose value decreases with
every iteration, but cannot decrease indefinitely as it is bounded from below.
This example also illustrates the lack of a direct correlation between the linearity of a program and that of its termination arguments.

Assumption (ii) is broken by the loop in Figure~\ref{fig:motivation.b}.
We prove that this loop terminates by finding that the non-linear function $R(x) = |x|$ is a possible ranking function.

Figure~\ref{fig:motivation.c} invalidates assumption (iii). 
This loop is  terminating for bit-vectors since \texttt{x}
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic as the loop condition stays always true for any initial \texttt{x} at least 1.

Assumption (iii) is broken again by the program in Figure~\ref{fig:motivation.d}. Every execution of the program can be partitioned into two phases: 
in the first phase $y$ increases until it is positive (in this phase $q$ may increase), whereas in the second $q$ decreases until the loop condition is violated. 
This example is taken from \cite{DBLP:conf/tacas/LeikeH14}, where the authors make use of a template for obtaining a ranking function that proceeds
through a fixed number of phases in the program execution. Each phase is ranked by a linear function, and ends when this function becomes non-positive.
However, when considering the bit-vector semantics, this program does not terminate as..\todo{change this example}

The loops in Figure~\ref{fig:motivation.e} and Figure~\ref{fig:motivation.f} break assumption (iv) as their terminating behaviour is different 
for floats and mathematical reals. 
%This assumption may lead to erroneous diagnosis of a program's terminating behaviour.
The former does not terminate for floats (if $x$ is sufficiently large, the rounding makes the subtraction negligible), but does for reals. Conversely, the latter 
doesn't terminate for reals, but does so for floats (due to rounding errors $x$ eventually becomes sufficiently small 
such that the closest representable number is 0.0). \todo{can we prove non-termination?}

@@

Figure~\ref{fig:motivation.g} is a linear program taken from \cite{DBLP:conf/tacas/CookSZ13}, 
where it is shown to not admit (without prior manipulation) a lexicographic linear function, but only a Ramsey-based termination argument,
which requires an expensive binary reachability analysis.
With our technique we can find a non-linear lexicographic ranking function $R(x) = |x|$ 
(a lexicographic non-linear ranking functions consists of lexicographically ordered components
of non-linear functions). 
%As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
%loop transition leads to a decrease with respect to the lexicographic
%ordering for this tuple. Therefore no function may increase unless a function of
%a lower index decreases. Additionally, at every step, there must be at least one
%function that decreases.

The loop in Figure~\ref{fig:motivation.i} is taken from SVCOMP'15  \footnote{http://sv-comp.sosy-lab.org/2015/index.php} termination benchmarks.
In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, this program admits a multiphase ranking function computed from a multiphase ranking template.
A multiphase ranking template is targeted at programs that go through a
finite number of phases in their execution. Each phase is ranked with
an affine-linear function and the phase is considered to be completed once this
function becomes non-positive.
However, in our setting this type of programs do not need a special treatment, as we can find a non-linear ranking function:
\begin{verbatim}
R(x, y, z) = (x < y, z)
\end{verbatim}

The loop in Figure~\ref{fig:motivation.h} illustrates conditional termination.
When proving program termination we are simultaneously solving two problems:
the search for a termination argument, and the search for a supporting invariant.
We iteratively construct a termination argument as a pair of a ranking function and supporting invariant through a single refinement loop.
For this loop, we find the ranking function $R(x) = x$ together with the supporting invariant $y=1$.

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0)
 x = (x - 1) & x;
\end{lstlisting}
\caption{}
 \label{fig:motivation.a}
\end{subfigure}%

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x != 0)
 x = -x / 2;
\end{lstlisting}
\caption{}
 \label{fig:motivation.b}
\end{subfigure}%

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}[language=C]
 while(x > 0) x++;
 \end{lstlisting}
\caption{}
 \label{fig:motivation.c}
\end{subfigure} \\

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
  while ( q >= 0 ) {
    q = q - y;
    y = y + 1;
  }
\end{lstlisting}
\caption{}
 \label{fig:motivation.d}
\end{subfigure}

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0.0) x -= 1.0;
\end{lstlisting}
\caption{}
 \label{fig:motivation.e}
\end{subfigure}

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0.0) x *= 0.5;
\end{lstlisting}
\caption{}
 \label{fig:motivation.f}
\end{subfigure} \\
\hline

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
  while (x != 0) {
    if (x > 0)
      x--;
    else
      x++;
  }
\end{lstlisting}
\caption{}
 \label{fig:motivation.g}
\end{subfigure}

@@

  while (x > 0) {
    x = x-y;
  }
\end{lstlisting}
\caption{}
 \label{fig:motivation.h}
\end{subfigure}

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0 && y > 0 && z > 0) {
    if (y > x) {
      y = z;
      x = nondet();
      z = x - 1;
    } else {
      z = z - 1;
      x = nondet();
      y = x - 1;
    }
}
\end{lstlisting}
\caption{}
 \label{fig:motivation.i}
\end{subfigure}

@@

\section{Soundness, Completeness and Complexity}
\begin{theorem}
 \label{thm:synth-sound}
 Algorithm~\ref{alg:abstract-refinement-code} is sound -- if it terminates with witness $P$ then
 $P \models \sigma$.
\end{theorem}

@@

\begin{theorem}
 \label{thm:synth-semicomplete}
 If the existential first-order theory used to express the specification $\sigma$ is decidable,
 algorithm~\ref{alg:abstract-refinement-code} is semi-complete -- if a program $P \models \sigma$
 exists then algorithm~\ref{alg:abstract-refinement-code} will terminate.  However if no program
 satisfies the specification, the algorithm may not terminate.  Furthermore if the domain of inputs $X$
 is finite, the algorithm is guaranteed to terminate.
\end{theorem}

\begin{proof}
 Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
 the program length $l$), it will eventually enumerate a program that meets the specification
 in whatever set of inputs are currently being tracked.  Since the first order theory is
 decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
 The set of correct programs is therefore recursively enumerable and algorithm~\ref{alg:abstract-refinement-code}
 enumerates this set, so it is semi-complete.

>>>>

One difficulty with a Ramsey-based approach is that validity of the termination argument can only be checked by considering the \emph{transitive closure}
of the program, rather than a single step.
As such, there is experimental evidence that most of the time is spent in reachability analysis \cite{DBLP:conf/pldi/CookPR06}, 
requiring the support of powerful safety checkers:
there is a tradeoff between the complexity of the termination arguments and that of checking their validity.

As Ramsey-based approaches are limited by the state of the art in safety checking, 
recent research shifts back to more complex termination arguments that are easier to check \cite{DBLP:conf/tacas/CookSZ13,DBLP:conf/cav/KroeningSTW10}.
%In \cite{DBLP:conf/tacas/CookSZ13}, Cook et al present an iterative termination proving procedure that searches for 
%lexicographic termination arguments, whereas Kroening et al. strengthen the termination argument such that it becomes a transitive relation \cite{DBLP:conf/cav/KroeningSTW10}.
%
%
Following the same trend, %of switching the complexity in termination checking back to the termination arguments, 
we investigate its extreme: \emph{unrestricted} termination arguments. 
This means that our ranking functions may involve non-linearity and lexicographic orders (we do not commit to any particular syntactic form, and do not use templates).

While the summary  in Figure~\ref{fig:handletable} supports the observation that the majority of existing termination analyses are designed for
linear programs and linear ranking functions, it also identifies another common assumption made by state of the art termination provers: 
that bit-vector semantics and integer semantics give rise to the same termination behaviour. Thus, most of the techniques treat
bit-vectors and floats as mathematical integers and reals respectively 
\cite{DBLP:conf/pldi/CookPR06,DBLP:conf/popl/Ben-AmramG13,DBLP:conf/vmcai/P04,DBLP:conf/atva/HeizmannHLP13,DBLP:conf/vmcai/BradleyMS05,DBLP:conf/cav/KroeningSTW10}.

By assuming bit-vector semantics to be identical to integer semantics,
these techniques ignore the wrap-around behaviour caused by overflows,
which can be unsound.
In Section~\ref{sec:motivation}, we show that integers and bit-vectors exhibit
incomparable behaviours with respect to termination, i.e.
programs that terminate for integers may \emph{not} terminate for bit-vectors and
vice-versa.
Thus, abstracting bit-vectors to integers may give rise to 
{\em unsound} and {\em incomplete} analyses.
%design decisions
%which is rather surprising given that bit-vectors and floats are ubiquitous in computer systems.

@@

We propose a general framework that %does not assume the existence of linear ranking functions and can 
uniformly computes
lexicographic, non-linear 
ranking functions supported by non-linear inductive invariants for loops with 
arbitrary guards and transitions over bit-vectors and floats.
It also computes non-linear recurrence sets, which comprise non-termination proofs.
Our encoding uniformly multiple loops and nested loops, without having to enumerate
lassos.
%Our technique is {\emph complete} and can handle programs with non-linear operations, e.g. logical and.
%--In our design, the termination problem becomes as hard as finding ranking functions, rather than as hard as
%checking the validity of a termination argument. 
%We combine the generation of ranking functions with the generation of invariants such that the necessary 
%supporting invariants and the ranking function are discovered at the same time.

For this purpose, we phrase the termination and non-termination problems as second order satisfaction problems.
We identify a fragment of second-order logic that is expressive enough to capture our formulation,
which we call the \emph{synthesis fragment}. Subsequently, we propose a method
for checking the satisfiability of a formula in the synthesis fragment by solving an isomorphic program synthesis problem (Section~\ref{sec:synthesis}).
Our method is sound and semi-complete, i.e. if a pair of a ranking function and a supporting invariant 
exist it will be found, but if the program does not terminate we may not find a recurrence set.  However
if the program's transition relation is deterministic, our method is complete.

With respect to the performance of our technique, we show that its runtime is dominated by the size of 
the shortest termination proof.
%While investigating the search space and the distribution of solutions (Section~\ref{}), we identify factors 
%influencing our method's performance and
%find that our technique  
This means that our procedure is not directly dependent on the structure of the analysed loop, e.g. number of variable, number of lassos, 
but on the Kolmogorov complexity of its minimal termination argument. 
We argue that the length of a program's termination proof is indicative of how easy to understand the program is, and thus 
programmers tend to write programs that have relatively short termination proofs. 
We empirically show that this claim holds in practice, causing our technique to perform well.

@@

Additionally, if the program under analysis has a non-deterministic transition relation
and is non-terminating, our procedure is not guaranteed to find a non-termination proof.

@@

Figure~\ref{fig:handletable} illustrates the most common assumptions made by existing termination analyses:
\begin{itemize}
\item[(i)] programs use only linear arithmetic.
\item[(ii)] terminating programs have termination arguments in linear arithmetic.
\item[(iii)] the semantics of bit-vectors and mathematical integers are equivalent.
\item[(iv)] the semantics of floats and mathematical reals are equivalent.
\end{itemize}

%In the previous sections, we have discussed some of the assumptions made by existent termination analyses (see Figure~\ref{fig:handletable}).
To show how these assumptions are broken by even simple programs, we draw the reader's attention to the programs in
Figure~\ref{fig:motivation} and their curious properties:

@@

%Figure~\ref{fig:handletable} shows that the most common assumption existent termination techniques make is that \emph{programs are linear}.

\begin{itemize}

\item Program (a) breaks assumption (i) as it makes use of the bit-wise $\&$ operator.
%As such, Figure~\ref{} denotes a program with non-linear operations. Given that  many of the existing techniques commit to linear programs, they cannot handle this situation, 
%although a linear ranking function does exist (see Figure~\ref{fig:handletable}). 
%In order to find a ranking function for this example, it is necessary to take into account
%the semantics of the bit-wise AND operator.
Our technique finds that an admissible ranking function is the linear function
$R(x) = x$, whose value decreases with
every iteration, but cannot decrease indefinitely as it is bounded from below.
This example also illustrates the lack of a direct correlation between the linearity of a program and that of its termination arguments.

\item Program (b) break assumption (ii), in that it has no linear ranking function.
We prove that this loop terminates by finding the non-linear ranking function $R(x) = |x|$.

\item Program (c) breaks assumption (iii).
This loop is  terminating for bit-vectors since $x$
will eventually over-flow and become negative. Conversely, the same program is non-terminating using integer
arithmetic since $x > 0 \rightarrow x+1 > 0$ for any integer $x$.

\item Program (d) also breaks assumption(iii), but ``the other way'' -- it terminates for integers but not for bit-vectors.
If each of the variables is stored in an unsigned $k$-bit word, the following configuration will
lead to an infinite loop:
$$M = 2^k - 1, N = 2^k - 1, i = M, j = N-1$$

\item Program (e) breaks assumption (iv) -- it terminates for reals but not for floats.
If $x$ is sufficiently large, rounding error will cause the subtraction to have no effect.

\item Program (f) breaks assumption (iv) ``the other way'' -- it terminates for floats but not for reals.
Eventually $x$ will become sufficiently small that the nearest representable number is 0.0, at which point
it will be rounded to 0.0 and the loop will terminate.
\end{itemize}

@@

\begin{itemize}
\item Program (g) is a linear program
which is shown in~\cite{DBLP:conf/tacas/CookSZ13} not to admit (without prior manipulation) a lexicographic linear function.
With our technique we can find the non-linear ranking function $R(x) = |x|$.
%As with linear lexicographic ranking functions, a state is mapped to a tuple of values such that the
%loop transition leads to a decrease with respect to the lexicographic
%ordering for this tuple. Therefore no function may increase unless a function of
%a lower index decreases. Additionally, at every step, there must be at least one
%function that decreases.

\item In the terminology of \cite{DBLP:conf/tacas/LeikeH14}, program (i) admits a multiphase ranking function computed from a multiphase ranking template.
A multiphase ranking template is targeted at programs that go through a
finite number of phases in their execution. Each phase is ranked with
an affine-linear function and the phase is considered to be completed once this
function becomes non-positive.

In our setting this type of programs does not need special treatment, as we can find a non-linear lexicographic ranking function
$R(x, y, z) = (x < y, z)$.  This termination argument is somewhat subtle, but can be understood as follows: the boolean quantity $x < y$
eventually decreases, that is to say once a state with $x \geq y$ is reached, $x$ never again becomes greater than $y$.  This means
that as soon as the ``else'' branch of the if statement is taken, it will continue to be taken in each subsequent iteration of the loop.
Meanwhile, if $x < y$ has not decreased (i.e. we have stayed in the same branch of the ``if'') then $z$ does decrease.  Since
a boolean only has two possible values, it cannot decrease indefinitely.  Since $z > 0$ is a conjunct of the loop guard,
$z$ cannot decrease indefinitely, and so $R$ proves that the loop is well founded.

As with all of the termination proofs presented in this paper, the ranking function above was found completely automatically.

\item Program (h) illustrates conditional termination.
When proving program termination we are simultaneously solving two problems:
the search for a termination argument, and the search for a supporting invariant~\cite{DBLP:conf/cav/BrockschmidtCF13}.
For this loop, we find the ranking function $R(x) = x$ together with the supporting invariant $y=1$.
\end{itemize}

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0) {
  x = (x - 1) & x;
}
\end{lstlisting}
\caption{Taken from~\cite{DBLP:conf/tacas/CookKRW10}.}
 \label{fig:motivation.a}
\end{subfigure}%

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x != 0) {
  x = -x / 2;
}
\end{lstlisting}
\caption{}
 \label{fig:motivation.b}
\end{subfigure}%

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}[language=C]
while(x > 0) {
  x++;
}
 \end{lstlisting}
\caption{}
 \label{fig:motivation.c}
\end{subfigure} \\

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (i < M || j < N) {
  i = i + 1;
  j = j + 1;
}
\end{lstlisting}
\caption{Taken from~\cite{DBLP:conf/sigsoft/Nori013}}
 \label{fig:motivation.d}
\end{subfigure}

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
float x;

while (x > 0.0) {
  x -= 1.0;
}
\end{lstlisting}
\caption{}
 \label{fig:motivation.e}
\end{subfigure}

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
float x;

while (x > 0.0) {
  x *= 0.5;
}
\end{lstlisting}
\caption{}
 \label{fig:motivation.f}
\end{subfigure} \\
\hline

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x != 0) {
  if (x > 0)
    x--;
  else
    x++;
}
\end{lstlisting}
\caption{Taken from \cite{DBLP:conf/tacas/CookSZ13}}
 \label{fig:motivation.g}
\end{subfigure}

@@

while (x > 0) {
  x = x - y;
}
\end{lstlisting}
\caption{}
 \label{fig:motivation.h}
\end{subfigure}

@@

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (x > 0 && y > 0 && z > 0) {
  if (y > x) {
    y = z;
    x = nondet();
    z = x - 1;
  } else {
    z = z - 1;
    x = nondet();
    y = x - 1;
  }
}
\end{lstlisting}
\caption{Taken from~\cite{BA:mcs}}
 \label{fig:motivation.i}
\end{subfigure}

\\

\begin{subfigure}[b]{0.3\textwidth}
\begin{lstlisting}
while (i < n) {
  j = 0;
  while (j <= i) {
    j = j + 1;
  }
  i = i + 1;
}
\end{lstlisting}
\caption{Taken from~\cite{DBLP:conf/cav/BrockschmidtCF13}}
\end{subfigure}

@@

\section{Soundness, Completeness and Complexity}
\begin{theorem}
 \label{thm:synth-sound}
 Algorithm~\ref{alg:cegis} is sound -- if it terminates with witness $P$ then
 $P \models \sigma$.
\end{theorem}

@@

\begin{theorem}
 \label{thm:synth-semicomplete}
 If the existential first-order theory used to express the specification $\sigma$ is decidable,
 algorithm~\ref{alg:cegis} is semi-complete -- if a program $P \models \sigma$
 exists then algorithm~\ref{alg:cegis} will terminate.  However if no program
 satisfies the specification, the algorithm may not terminate.  Furthermore if the domain of inputs $X$
 is finite, the algorithm is guaranteed to terminate.
\end{theorem}

\begin{proof}
 Since the {\sc ExplicitSearch} routine enumerates all programs (as can be seen by induction on
 the program length $l$), it will eventually enumerate a program that meets the specification
 in whatever set of inputs are currently being tracked.  Since the first order theory is
 decidable, the query in {\sc Verif} will succeed for this program, causing the algorithm to terminate.
 The set of correct programs is therefore recursively enumerable and algorithm~\ref{alg:cegis}
 enumerates this set, so it is semi-complete.
